---
title: "Early Infection - Ileum - Gene"
author: "Jared J. Homola"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---
This is an analysis of differential expression at the gene level between supershedder groupings based on shedding rates measured on the day of sacrifice for infected mallard ileum samples.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r message=FALSE, warning=FALSE}
## Load packages and data
library(limma)
library(edgeR)
library(Glimma)
library(gplots)
library(ggrepel)
library(RColorBrewer)
library(gridExtra)
library(kableExtra)
library(org.Gg.eg.db)
library(tidyverse)

## Load data
cnt1 <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/ileumCounts.txt", delim = "\t") %>% distinct(gene, .keep_all = TRUE)
covars <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/MALL70_SSgroup_RAW_7.7.20.csv", delim = ",")
annot <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/annotationGenesIleum110120.txt", delim = "\t")
```


### Prep data for analysis
```{r, warning = FALSE, message = FALSE}
cnt <- cnt1 %>%
  as_tibble() %>%
  select(-gene)

newNames.ileum <- names(cnt) %>%
  as_tibble() %>%
  mutate(value = gsub("Iluem_", "", .$value)) %>% 
  separate(value, into = c(NA, "pt1", "pt2")) %>%
  mutate(pt1 = na_if(pt1, "MALL"),
         pt2 = na_if(pt2, "ileum")) %>%
  mutate(bird = ifelse(is.na(pt1), pt2, pt1)) %>%
  mutate(bird = as.numeric(bird),
    bird = formatC(bird, width = 2, format = "d", flag = "0")) %>%
  mutate(newName = paste0("ileum_", bird))

colnames(cnt) <- newNames.ileum$newName

cnt <- cnt[,order(colnames(cnt))] %>%
  mutate(gene = cnt1$gene) %>%
  column_to_rownames("gene") %>%
  as_tibble(rownames = NA)

### Covars
covars <- covars %>%
  mutate(bird = as.numeric(bird)) %>%
  mutate(bird = formatC(bird, width = 2, format = "d", flag = "0")) %>%
  arrange(bird) %>%
  mutate(bird = as.factor(bird)) %>%
  mutate(group = str_remove(group, "-"))

covars.ileum <- covars %>%
  filter(bird %in% newNames.ileum$bird)

cnt <- cnt %>% select(-ileum_01, -ileum_72)

covars.ileum <- covars.ileum %>% filter(bird != "01",
                                        bird != "72")
```

### Set up model effects
```{r, warning = FALSE, message = FALSE}
bird <- as.factor(covars.ileum$bird)
sex <- as.factor(covars.ileum$sex)
age <- as.numeric(covars.ileum$age)
weight <- covars.ileum$wt_55
group <- recode(covars.ileum$group, C1 = "Ctl", C29 = "Ctl") %>% 
  as.factor()
pool <- as.factor(covars.ileum$Ileum_pool)
```

### Prep data for analysis
Here we standardize expression levels across samples of varying depth by using the counts per million (CPM) and trimmed mean of M-values (TMM) methods. We also remove any genes that are expressed at few than 0.5 CPM in at least 25% of individuals.
```{r, warning = FALSE, message = FALSE}
#Convert to DGEList object
dge <- DGEList(counts=cnt)
dge$genes <- annot[match(rownames(dge$counts), annot$ensembl_gene_id),]

#CPM and log-CPM
cpm <- cpm(dge)
lcpm <- cpm(dge, log = TRUE)

#### Retain only genes expressed at >0.5 CPM in at least 25% of individuals
table(rowSums(dge$counts==0) == length(cnt[1,]))
keep.exprs <- rowSums(cpm>0.5) >= length(cnt[1,])/4
sum(keep.exprs)

dge <- dge[keep.exprs,, keep.lib.sizes=FALSE]
table(rowSums(dge$counts==0) == length(cnt[1,]))
dim(dge)

#TMM normalization
dge <- calcNormFactors(dge, method = "TMM")
```

### Subset data
Following normalization, we subset the data based on infection groups.
```{r, warning = FALSE, message = FALSE}
#Assign groups in DGElist
dge$samples$group <- group

### Divide out I1 and I3
dge.I1 <- dge[,grep('\\bI1\\b', dge$samples$group)]
dge.I2 <- dge[,grep('\\bI2\\b', dge$samples$group)]

covars.I1 <- covars %>% filter(group == "I1")
covars.I2 <- covars %>% filter(group == "I2")
```


# Differential expression analysis
### Establish design & contrast matrices
```{r}
## Establish design matrix: I1
design.I1 = model.matrix(~ 0 +
                        factor(covars.I1$SSgroup.virus.sac) +
                        covars.I1$age +
                        covars.I1$sex + 
                        covars.I1$wt_55 + 
                        covars.I1$Bursa_pool)

colnames(design.I1) <- c("HIGH", "LOW", "MODERATE",
                         "age", "sexM", "weight", "poolMSU2",
                         "poolMSU3", "poolMSU4", "poolMSU5")


contr.matrix.I1 = makeContrasts(
  LvM = LOW - MODERATE,
  MvH = MODERATE - HIGH,
  LvH = LOW - HIGH,
  levels = colnames(design.I1))

## Mean-variance trend plots
v.I1 <- voomWithQualityWeights(dge.I1, design.I1, plot = FALSE)

#Fitting models
vfit.I1 <- lmFit(v.I1, design.I1) 
vfit.I1 <- contrasts.fit(vfit.I1, contrasts = contr.matrix.I1)
tfit.I1 <- treat(vfit.I1, lfc = 1.0)

## Identify DE genes
dt.I1 <- decideTests(tfit.I1, p.value = 0.1, adjust.method = "fdr")
dt.tib.I1 <- as_tibble(dt.I1, rownames = NA) %>% 
  rownames_to_column("gene") %>% 
  mutate_at(vars(starts_with("L")), as.numeric) %>% 
  mutate_at(vars(starts_with("M")), as.numeric) %>% 
  rename(LvM.I1 = LvM,
         MvH.I1 = MvH,
         LvH.I1 = LvH)


## Establish design matrix: I2
design.I2 = model.matrix(
  ~ 0 +
    factor(covars.I2$SSgroup.virus.sac) +
    covars.I2$age +
    covars.I2$sex +
    covars.I2$wt_55 #+
    #covars.I2$Bursa_pool
)

colnames(design.I2) <- c(
  "HIGH",
  "LOW",
  "MODERATE",
  "age",
  "sexM",
  "weight"#,
  #"poolMSU2",
  #"poolMSU3",
  #"poolMSU4",
  #"poolMSU5"
)

contr.matrix.I2 = makeContrasts(
  LvM = LOW - MODERATE,
  MvH = MODERATE - HIGH,
  LvH = LOW - HIGH,
  levels = colnames(design.I2))

## Mean-variance trend plots
v.I2 <- voomWithQualityWeights(dge.I2, design.I2, plot = FALSE)

#Fitting models
vfit.I2 <- lmFit(v.I2, design.I2) 
vfit.I2 <- contrasts.fit(vfit.I2, contrasts = contr.matrix.I2)
tfit.I2 <- treat(vfit.I2, lfc = 1.0)

## Identify DE genes
dt.I2 <- decideTests(tfit.I2, p.value = 0.1, adjust.method = "fdr")
dt.tib.I2 <- as_tibble(dt.I2, rownames = NA) %>% 
  rownames_to_column("gene") %>% 
  mutate_at(vars(starts_with("L")), as.numeric) %>% 
  mutate_at(vars(starts_with("M")), as.numeric) %>% 
  rename(LvM.I2 = LvM,
         MvH.I2 = MvH,
         LvH.I2 = LvH)

## Bring it all together
dt.tib <- left_join(dt.tib.I1, dt.tib.I2) %>% 
  mutate_at(vars(starts_with("L")), as.numeric) %>% 
  mutate_at(vars(starts_with("M")), as.numeric) %>% 
  filter_at(vars(LvM.I1:LvH.I2), any_vars(. != 0))
```

### Mean-variance trend and sample weight plots
```{r}
v <- voomWithQualityWeights(dge, design, plot = TRUE)
```

### Fitting the model
```{r}
vfit <- lmFit(v, design) 
vfit <- contrasts.fit(vfit, contrasts = contr.matrix)
tfit <- treat(vfit, lfc = 1)
plotSA(tfit)
```

### Count DE genes
For a gene to be considered differentially expressed, we require a p-value of 0.1 with a false discovery rate correction and a log fold count difference of 1.
```{r}
dt <- decideTests(tfit, p.value = 0.1, adjust.method = "fdr")
summary(dt) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F)

dt.tib <- as_tibble(dt, rownames = NA) %>% 
  rownames_to_column("gene") %>% 
  mutate_at(vars(starts_with("C")), as.numeric) %>% 
  mutate_at(vars(starts_with("I")), as.numeric) %>% 
  filter_at(vars(CtlvI1:I15vI29), any_vars(. != 0))
```

### Histogram of P-values
```{r}
hist(tfit$p.value, main = "P-values comparing groups (eBayes)")
```


### Volcano plot
```{r, warning = FALSE, message = FALSE, fig.height = 20, fig.width = 7, fig.align = "center"}
tmp1 <- topTreat(tfit, coef = 1, n = Inf)
results1 <- mutate(tmp1, sig=ifelse(tmp1$adj.P.Val<0.1, "Sig", "Not Sig"))
p1 <- ggplot(results1, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 1") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp2 <- topTreat(tfit, coef = 2, n = Inf)
results2 <- mutate(tmp2, sig=ifelse(tmp2$adj.P.Val<0.1, "Sig", "Not Sig"))
p2 <- ggplot(results2, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 2") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp3 <- topTreat(tfit, coef = 3, n = Inf)
results3 <- mutate(tmp3, sig=ifelse(tmp3$adj.P.Val<0.1, "Sig", "Not Sig"))
p3 <- ggplot(results3, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 5") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp4 <- topTreat(tfit, coef = 4, n = Inf)
results4 <- mutate(tmp4, sig=ifelse(tmp4$adj.P.Val<0.1, "Sig", "Not Sig"))
p4 <- ggplot(results4, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 15") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp5 <- topTreat(tfit, coef = 5, n = Inf)
results5 <- mutate(tmp5, sig=ifelse(tmp5$adj.P.Val<0.1, "Sig", "Not Sig"))
p5 <- ggplot(results5, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 29") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp6 <- topTreat(tfit, coef = 6, n = Inf)
results6 <- mutate(tmp6, sig=ifelse(tmp6$adj.P.Val<0.1, "Sig", "Not Sig"))
p6 <- ggplot(results6, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 2") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")


tmp7 <- topTreat(tfit, coef = 7, n = Inf)
results7 <- mutate(tmp7, sig=ifelse(tmp7$adj.P.Val<0.1, "Sig", "Not Sig"))
p7 <- ggplot(results7, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 5") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp8 <- topTreat(tfit, coef = 8, n = Inf)
results8 <- mutate(tmp8, sig=ifelse(tmp8$adj.P.Val<0.1, "Sig", "Not Sig"))
p8 <- ggplot(results8, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 15") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp9 <- topTreat(tfit, coef = 9, n = Inf)
results9 <- mutate(tmp9, sig=ifelse(tmp9$adj.P.Val<0.1, "Sig", "Not Sig"))
p9 <- ggplot(results9, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 29") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp10 <- topTreat(tfit, coef = 10, n = Inf)
results10 <- mutate(tmp10, sig=ifelse(tmp10$adj.P.Val<0.1, "Sig", "Not Sig"))
p10 <- ggplot(results10, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 2 vs. Infected, Day 5") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp11 <- topTreat(tfit, coef = 11, n = Inf)
results11 <- mutate(tmp11, sig=ifelse(tmp11$adj.P.Val<0.1, "Sig", "Not Sig"))
p11 <- ggplot(results11, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 2 vs. Infected, Day 15") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp12 <- topTreat(tfit, coef = 12, n = Inf)
results12 <- mutate(tmp12, sig=ifelse(tmp12$adj.P.Val<0.1, "Sig", "Not Sig"))
p12 <- ggplot(results12, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 2 vs. Infected, Day 29") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp13 <- topTreat(tfit, coef = 13, n = Inf)
results13 <- mutate(tmp13, sig=ifelse(tmp13$adj.P.Val<0.1, "Sig", "Not Sig"))
p13 <- ggplot(results13, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 5 vs. Infected, Day 15") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp14 <- topTreat(tfit, coef = 14, n = Inf)
results14 <- mutate(tmp14, sig=ifelse(tmp14$adj.P.Val<0.1, "Sig", "Not Sig"))
p14 <- ggplot(results14, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 5 vs. Infected, Day 29") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp15 <- topTreat(tfit, coef = 15, n = Inf)
results15 <- mutate(tmp15, sig=ifelse(tmp15$adj.P.Val<0.1, "Sig", "Not Sig"))
p15 <- ggplot(results15, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 15 vs. Infected, Day 29") +
  ylab("-Log10(Adjusted p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")


grid.arrange(p1, p2, p3, p4, p5, p6, p7,
             p8, p9, p10, p11, p12, p13, p14, p15, nrow = 8)
```

