---
title: "Infected and Control - Bursa - Transcript"
author: "Jared J. Homola"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---
This is an analysis of differential expression at the transcript level between treatment groups for control and infected mallard bursa samples.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
## Load packages and data
library(limma)
library(edgeR)
library(Glimma)
library(gplots)
library(ggrepel)
library(RColorBrewer)
library(gridExtra)
library(kableExtra)
library(org.Gg.eg.db)
library(tidyverse)

## Load data
cnt <- read.table("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/mallardLPAIV/extData/all.isoforms.results.csv", header = TRUE) %>% as_tibble(rownames = NA)
covars <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/mallardLPAIV/extData/MALL70_SSgroup_RAW_11.5.20.csv", delim = ",")
annot <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/mallardLPAIV/extData/annotationTrans103120.txt", delim = "\t")
```


### Prep data for analysis
```{r, warning = FALSE, message = FALSE}
newNames <- names(cnt) %>%
  as_tibble() %>%
  separate(value, into = c("pt1", "pt2", "pt3")) %>%
  select(pt2, pt3) %>%
  mutate(pt2 = as.numeric(pt2),
         pt2 = formatC(pt2, width = 2, format = "d", flag = "0")) %>%
  mutate(newName = paste0("bursa_", pt2))

colnames(cnt) <- newNames$newName

cnt <- cnt %>% 
  select(order(colnames(cnt)))

covars <- covars %>%
  mutate(bird = as.numeric(bird)) %>%
  mutate(bird = formatC(bird, width = 2, format = "d", flag = "0")) %>%
  arrange(bird) %>%
  mutate(bird = as.factor(bird)) %>%
  mutate(group = str_remove(group, "-"))

## Remove birds with issues
cnt <- cnt %>% select(-bursa_01, -bursa_72)

covars <- covars %>% filter(bird != "01", bird != "72")
```

### Set up model effects
```{r, warning = FALSE, message = FALSE}
bird <- as.factor(covars$bird)
sex <- as.factor(covars$sex)
age <- as.numeric(covars$age)
group <- recode(covars$group, C1 = "Ctl", C29 = "Ctl") %>% 
  as.factor()
pool <- as.factor(covars$Bursa_pool)
```

### Prep data for analysis
Here we standardize expression levels across samples of varying depth by using the counts per million (CPM) and trimmed mean of M-values (TMM) methods. We also remove any genes that are expressed at few than 0.5 CPM in at least 25% of individuals.
```{r, warning = FALSE, message = FALSE}
#Convert to DGEList object
dge <- DGEList(counts=cnt)
dge$genes <- annot[match(rownames(dge$counts), annot$ensembl_transcript_id),]

#CPM and log-CPM
cpm <- cpm(dge)
lcpm <- cpm(dge, log = TRUE)

#### Retain only genes expressed at >0.5 CPM in at least 25% of individuals
table(rowSums(dge$counts==0) == length(cnt[1,]))
keep.exprs <- rowSums(cpm>0.5) >= length(cnt[1,])/4
sum(keep.exprs)

dge <- dge[keep.exprs,, keep.lib.sizes=FALSE]
table(rowSums(dge$counts==0) == length(cnt[1,]))
dim(dge)

#TMM normalization
dge <- calcNormFactors(dge, method = "TMM")
```


# Differential expression analysis
### Establish design & contrast matrices
```{r}
## Establish design matrix
design = model.matrix(~ 0 +
                        group +
                        age +
                        sex + 
                        pool)

colnames(design) <- gsub("group", "", colnames(design))

contr.matrix = makeContrasts(
  CtlvI1 = Ctl - I1,
  CtlvI2 = Ctl - I2,
  CtlvI5 = Ctl - I5,
  CtlvI15 = Ctl - I15,
  CtlvI29 = Ctl - I29,
  I1vI2 = I1 - I2,
  I1vI5 = I1 - I5,
  I1vI15 = I1 - I15,
  I1vI29 = I1 - I29,
  I2vI5 = I2 - I5,
  I2vI15 = I2 - I15,
  I2vI29 = I2 - I29,
  I5vI15 = I5 - I15,
  I5vI29 = I5 - I29,
  I15vI29 = I15 - I29,
  levels = colnames(design))

### Mean-variance trend and sample weight plots
v <- voomWithQualityWeights(dge, design, plot = TRUE)

### Fitting the model
vfit <- lmFit(v, design) 
vfit <- contrasts.fit(vfit, contrasts = contr.matrix)
tfit <- treat(vfit, lfc = 1)
plotSA(tfit)
```

### Count DE genes
For a gene to be considered differentially expressed, we require a p-value of 0.1 with a false discovery rate correction and a log fold count difference of 1.
```{r}
dt <- decideTests(tfit, p.value = 0.1, adjust.method = "fdr")
summary(dt) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F)

dt.tib <- as_tibble(dt, rownames = NA) %>% 
  rownames_to_column("gene") %>% 
  mutate_at(vars(starts_with("C")), as.numeric) %>% 
  mutate_at(vars(starts_with("I")), as.numeric) %>% 
  filter_at(vars(CtlvI1:I15vI29), any_vars(. != 0))
```


### Volcano plot
```{r, warning = FALSE, message = FALSE, fig.height = 20, fig.width = 7, fig.align = "center"}
tmp1 <- topTreat(tfit, coef = 1, n = Inf)
results1 <- mutate(tmp1, sig=ifelse(tmp1$adj.P.Val<0.1, "Sig", "Not Sig"))
p1 <- ggplot(results1, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 1") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp2 <- topTreat(tfit, coef = 2, n = Inf)
results2 <- mutate(tmp2, sig=ifelse(tmp2$adj.P.Val<0.1, "Sig", "Not Sig"))
p2 <- ggplot(results2, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 2") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp3 <- topTreat(tfit, coef = 3, n = Inf)
results3 <- mutate(tmp3, sig=ifelse(tmp3$adj.P.Val<0.1, "Sig", "Not Sig"))
p3 <- ggplot(results3, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 5") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp4 <- topTreat(tfit, coef = 4, n = Inf)
results4 <- mutate(tmp4, sig=ifelse(tmp4$adj.P.Val<0.1, "Sig", "Not Sig"))
p4 <- ggplot(results4, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 15") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp5 <- topTreat(tfit, coef = 5, n = Inf)
results5 <- mutate(tmp5, sig=ifelse(tmp5$adj.P.Val<0.1, "Sig", "Not Sig"))
p5 <- ggplot(results5, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Control vs. Infected, Day 29") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp6 <- topTreat(tfit, coef = 6, n = Inf)
results6 <- mutate(tmp6, sig=ifelse(tmp6$adj.P.Val<0.1, "Sig", "Not Sig"))
p6 <- ggplot(results6, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 2") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")


tmp7 <- topTreat(tfit, coef = 7, n = Inf)
results7 <- mutate(tmp7, sig=ifelse(tmp7$adj.P.Val<0.1, "Sig", "Not Sig"))
p7 <- ggplot(results7, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 5") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp8 <- topTreat(tfit, coef = 8, n = Inf)
results8 <- mutate(tmp8, sig=ifelse(tmp8$adj.P.Val<0.1, "Sig", "Not Sig"))
p8 <- ggplot(results8, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 15") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp9 <- topTreat(tfit, coef = 9, n = Inf)
results9 <- mutate(tmp9, sig=ifelse(tmp9$adj.P.Val<0.1, "Sig", "Not Sig"))
p9 <- ggplot(results9, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 1 vs. Infected, Day 29") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp10 <- topTreat(tfit, coef = 10, n = Inf)
results10 <- mutate(tmp10, sig=ifelse(tmp10$adj.P.Val<0.1, "Sig", "Not Sig"))
p10 <- ggplot(results10, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 2 vs. Infected, Day 5") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp11 <- topTreat(tfit, coef = 11, n = Inf)
results11 <- mutate(tmp11, sig=ifelse(tmp11$adj.P.Val<0.1, "Sig", "Not Sig"))
p11 <- ggplot(results11, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 2 vs. Infected, Day 15") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp12 <- topTreat(tfit, coef = 12, n = Inf)
results12 <- mutate(tmp12, sig=ifelse(tmp12$adj.P.Val<0.1, "Sig", "Not Sig"))
p12 <- ggplot(results12, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 2 vs. Infected, Day 29") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp13 <- topTreat(tfit, coef = 13, n = Inf)
results13 <- mutate(tmp13, sig=ifelse(tmp13$adj.P.Val<0.1, "Sig", "Not Sig"))
p13 <- ggplot(results13, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 5 vs. Infected, Day 15") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp14 <- topTreat(tfit, coef = 14, n = Inf)
results14 <- mutate(tmp14, sig=ifelse(tmp14$adj.P.Val<0.1, "Sig", "Not Sig"))
p14 <- ggplot(results14, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 5 vs. Infected, Day 29") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")

tmp15 <- topTreat(tfit, coef = 15, n = Inf)
results15 <- mutate(tmp15, sig=ifelse(tmp15$adj.P.Val<0.1, "Sig", "Not Sig"))
p15 <- ggplot(results15, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Infected, Day 15 vs. Infected, Day 29") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold count)") +
  theme_bw() +
  theme(legend.position = "none")


grid.arrange(p1, p2, p3, p4, p5, p6, p7,
             p8, p9, p10, p11, p12, p13, p14, p15, nrow = 8)
```

### Boxplots for differentially expressed genes
```{r message=FALSE, warning=FALSE}
lcpm.DE <- lcpm %>%
  as_tibble(rownames = NA) %>% 
  rownames_to_column("identifier") %>% 
  filter(identifier %in% dt.tib$gene) %>%
  pivot_longer(cols = contains("_"),
               names_to = "sample",
               values_to = "lcpm") %>%
  separate(sample, into = c("tissue", "bird")) %>%
  mutate(bird = as.factor(bird)) %>%
  left_join(covars, by = "bird") %>%
  select(identifier, bird, lcpm, group) %>% 
  mutate(group = recode(group, C29 = "Control"),
         group = recode(group, C1 = "Control"),
         tmpID = ifelse(group == "Control", 'Control', 'Infected'))


aprioriPlotting <- function(target, ...) {
   annot.target <- annot %>%
    select(
      ensembl_transcript_id,
      hgnc_symbol) %>%
    filter(ensembl_transcript_id == target) %>% 
    mutate(hgnc_symbol = replace_na(hgnc_symbol, "."))


  plot <- lcpm.DE %>%
    filter(identifier == target) %>%
    mutate(group = fct_relevel(group, "Control", "I1", "I2", "I5", "I15", "I29")) %>%
    ggplot(aes(x = group, y = lcpm)) +
    facet_grid(. ~ tmpID, scales = "free", space = "free") +
    ylab("Log2(Counts per million)") +
    xlab("Group") +
    geom_point(position = position_dodge(width=0.75), show.legend = FALSE) +
    geom_boxplot(alpha = 0.5) +
    geom_label_repel(aes(label = bird, fill = NULL), show.legend=FALSE) +
    theme_classic() +
    labs(title= paste0(target, " - ", annot.target[1,2])) +
    theme(legend.title = element_blank())
    
  print(plot)
}

for(target in sort(unique(lcpm.DE$identifier))) {
  aprioriPlotting(target)
}
```

### Annotations for differentially expressed genes
ns denotes non-significant genes for each comparison and numerical values are the log(fold count) difference
```{r message=FALSE, warning=FALSE, echo=FALSE}
tmp1a <- tmp1 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "CtlvI1")
tmp2a <- tmp2 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "CtlvI2")
tmp3a <- tmp3 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "CtlvI5")
tmp4a <- tmp4 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "CtlvI15")
tmp5a <- tmp5 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "CtlvI29")
tmp6a <- tmp6 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I1vI2")
tmp7a <- tmp7 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I1vI5")
tmp8a <- tmp8 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I1vI15")
tmp9a <- tmp9 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I1vI29")
tmp10a <- tmp10 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I2vI5")
tmp11a <- tmp11 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I2vI15")
tmp12a <- tmp12 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I2vI29")
tmp13a <- tmp13 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I5vI15")
tmp14a <- tmp14 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I5vI29")
tmp15a <- tmp15 %>% select(ensembl_transcript_id, adj.P.Val, logFC, hgnc_symbol, description) %>% mutate(comp = "I15vI29")

bind_rows(tmp1a, tmp2a, tmp3a, tmp4a, tmp5a, tmp6a, tmp7a, tmp8a, tmp9a, tmp10a, tmp11a, tmp12a, tmp13a, tmp14a, tmp15a) %>% 
  select(-description) %>% 
  filter(ensembl_transcript_id %in% dt.tib$gene) %>% 
  mutate(logFoldCount = ifelse(adj.P.Val < 0.10, round(logFC, 2), "ns")) %>% 
  select(-adj.P.Val, -logFC) %>% 
  pivot_wider(names_from = comp, values_from = logFoldCount) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

#### Gene functions
```{r message=FALSE, warning=FALSE}
bind_rows(tmp1a, tmp2a, tmp3a, tmp4a, tmp5a, tmp6a, tmp7a, tmp8a, tmp9a, tmp10a, tmp11a, tmp12a, tmp13a, tmp14a, tmp15a) %>% 
  select(-comp, -adj.P.Val, -logFC) %>% 
  filter(ensembl_transcript_id %in% dt.tib$gene) %>% 
  distinct(.) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```
