---
title: "Late stage infection - Ileum - Gene"
author: "Jared J. Homola"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---
This is an analysis of differential expression at the gene level between shedding rate groupings based on average shedding rates for infected mallard ileum samples.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r message=FALSE, warning=FALSE}
## Load packages and data
library(limma)
library(edgeR)
library(Glimma)
library(ggrepel)
library(gplots)
library(RColorBrewer)
library(gridExtra)
library(kableExtra)
library(tidyverse)

## Load data
cnt1 <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/mallardLPAIV/extData/ileumCounts.txt", delim = "\t") %>% distinct(gene, .keep_all = TRUE)
covars <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/mallardLPAIV/extData/MALL70_SSgroup_RAW_11.5.20.csv", delim = ",")
annot <- read_delim("G:/Shared drives/RNA Seq Supershedder Project/MALL DE manuscript/mallardLPAIV/extData/annotationGenesIleum110120.txt", delim = "\t")
```

### Prep data for analysis
```{r, warning = FALSE, message = FALSE}
cnt <- cnt1 %>%
  as_tibble() %>%
  select(-gene)

newNames.ileum <- names(cnt) %>%
  as_tibble() %>%
  mutate(value = gsub("Iluem_", "", .$value)) %>% 
  separate(value, into = c(NA, "pt1", "pt2")) %>%
  mutate(pt1 = na_if(pt1, "MALL"),
         pt2 = na_if(pt2, "ileum")) %>%
  mutate(bird = ifelse(is.na(pt1), pt2, pt1)) %>%
  mutate(bird = as.numeric(bird),
    bird = formatC(bird, width = 2, format = "d", flag = "0")) %>%
  mutate(newName = paste0("ileum_", bird))

colnames(cnt) <- newNames.ileum$newName

cnt <- cnt[,order(colnames(cnt))] %>%
  mutate(gene = cnt1$gene) %>%
  column_to_rownames("gene") %>%
  as_tibble(rownames = NA)

### Covars
covars <- covars %>%
  mutate(bird = as.numeric(bird)) %>%
  mutate(bird = formatC(bird, width = 2, format = "d", flag = "0")) %>%
  arrange(bird) %>%
  mutate(bird = as.factor(bird)) %>%
  mutate(group = str_remove(group, "-"))

covars.ileum <- covars %>%
  filter(bird %in% newNames.ileum$bird)

cnt <- cnt %>% select(-ileum_01, -ileum_72)

covars.ileum <- covars.ileum %>% filter(bird != "01",
                                        bird != "72")
```

### Prep data for analysis
Here we standardize expression levels across samples of varying depth by using the counts per million (CPM) and trimmed mean of M-values (TMM) methods. We also remove any genes that are expressed at few than 0.5 CPM in at least 25% of individuals.
```{r, warning = FALSE, message = FALSE}
#Convert to DGEList object
dge <- DGEList(counts=cnt)
dge$genes <- annot[match(rownames(dge$counts), annot$ensembl_gene_id),]

#CPM and log-CPM
cpm <- cpm(dge)
lcpm <- cpm(dge, log = TRUE)

#### Retain only genes expressed at >0.5 CPM in at least 25% of individuals
table(rowSums(dge$counts==0) == length(cnt[1,]))
keep.exprs <- rowSums(cpm>0.5) >= length(cnt[1,])/4
sum(keep.exprs)

dge <- dge[keep.exprs,, keep.lib.sizes=FALSE]
table(rowSums(dge$counts==0) == length(cnt[1,]))
dim(dge)
```


### Subset data
Following normalization, we subset the data based on infection groups.
```{r, warning = FALSE, message = FALSE}
#Assign groups in DGElist
group <- recode(covars.ileum$group, C1 = "Ctl", C29 = "Ctl") %>% 
  as.factor()
dge$samples$group <- group

### Divide out I5
dge.I5 <- dge[,grep('\\bI5\\b', dge$samples$group)]
dge.I5 <- calcNormFactors(dge.I5, method = "TMM")

covars.I5 <- covars.ileum %>% filter(group == "I5")
```

# Differential expression analysis
```{r, warning = FALSE, message = FALSE,}
## Establish design matrix: I5
design.I5 = model.matrix(~ 0 +
                        factor(covars.I5$SSgroup.virus.avg) +
                        covars.I5$sex + 
                        covars.I5$Ileum_pool)

colnames(design.I5) <- c("HIGH", "LOW", "MODERATE",
                         "sexM", "poolMSU_6",
                         "poolUMN")

contr.matrix.I5 = makeContrasts(
  LvM = LOW - MODERATE,
  MvH = MODERATE - HIGH,
  LvH = LOW - HIGH,
  levels = colnames(design.I5))

## Mean-variance trend plots
v.I5 <- voomWithQualityWeights(dge.I5, design.I5, plot = FALSE)

#Fitting models
vfit.I5 <- lmFit(v.I5, design.I5) 
vfit.I5 <- contrasts.fit(vfit.I5, contrasts = contr.matrix.I5)
tfit.I5 <- treat(vfit.I5, lfc = 0.5)

## Identify DE genes
dt.I5 <- decideTests(tfit.I5, p.value = 0.1, adjust.method = "fdr")
dt.tib.I5 <- as_tibble(dt.I5, rownames = NA) %>% 
  rownames_to_column("gene") %>% 
  mutate_at(vars(starts_with("L")), as.numeric) %>% 
  mutate_at(vars(starts_with("M")), as.numeric) %>% 
  rename(LvM.I5 = LvM,
         MvH.I5 = MvH,
         LvH.I5 = LvH)

dt.tib <- dt.tib.I5 %>% 
  mutate_at(vars(starts_with("L")), as.numeric) %>% 
  mutate_at(vars(starts_with("M")), as.numeric) %>% 
  filter_at(vars(LvM.I5:LvH.I5), any_vars(. != 0))
```

### Numbers of DE genes
```{r}
allResults <- as.data.frame(summary(dt.I5)) %>% 
  filter(Var1 != "NotSig")

kable(allResults) %>%
  kable_styling("striped", full_width = F)
```

### Volcano plot
```{r, warning = FALSE, message = FALSE, fig.height = 10, fig.width = 10, fig.align = "center"}
tmp1 <- topTreat(tfit.I5, coef = 1, n = Inf)
results1 <- mutate(tmp1, sig=ifelse(tmp1$adj.P.Val<0.1, "Sig", "Not Sig"))
p1 <- ggplot(results1, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Low vs. Moderate shedding") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold change)") +
  theme_bw() +
  theme(legend.position = "none")

tmp2 <- topTreat(tfit.I5, coef = 2, n = Inf)
results2 <- mutate(tmp2, sig=ifelse(tmp2$adj.P.Val<0.1, "Sig", "Not Sig"))
p2 <- ggplot(results2, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Moderate vs. High shedding") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold change)") +
  theme_bw() +
  theme(legend.position = "none")

tmp3 <- topTreat(tfit.I5, coef = 3, n = Inf)
results3 <- mutate(tmp3, sig=ifelse(tmp3$adj.P.Val<0.1, "Sig", "Not Sig"))
p3 <- ggplot(results3, aes(logFC, -log10(P.Value))) +
  geom_point(aes(col = sig)) +
  scale_color_manual(values=c("black", "red")) +
  ggtitle("Low vs. High shedding") +
  ylab("-Log10(p-value)") +
  xlab("Log(Fold change)") +
  theme_bw() +
  theme(legend.position = "none")

grid.arrange(p1, p2, p3, nrow = 1)
```


### Boxplots for differentially expressed genes
```{r message=FALSE, warning=FALSE}
lcpm.DE <- lcpm %>%
  as_tibble(rownames = NA) %>% 
  rownames_to_column("identifier") %>% 
  filter(identifier %in% dt.tib$gene) %>%
  pivot_longer(cols = contains("_"),
               names_to = "sample",
               values_to = "lcpm") %>%
  separate(sample, into = c("tissue", "bird")) %>%
  mutate(bird = as.factor(bird)) %>%
  left_join(covars, by = "bird") %>%
  filter(group == "C1" | group == "C14" | group == "I5") %>% 
  select(identifier, bird, lcpm, SSgroup.virus.avg, group) %>% 
  mutate(group = recode(group, C29 = "Control"),
         group = recode(group, C1 = "Control"),
         tmpID = ifelse(group == "Control", 'Control', 'Infected'))

aprioriPlotting <- function(target, ...) {
   annot.target <- annot %>%
    select(
      ensembl_gene_id,
      hgnc_symbol) %>%
    filter(ensembl_gene_id == target) %>% 
    mutate(hgnc_symbol = replace_na(hgnc_symbol, "."))

  plot <- lcpm.DE %>%
    filter(identifier == target) %>%
    mutate(SSgroup.virus.avg = fct_relevel(SSgroup.virus.avg, "LOW", "MODERATE", "HIGH")) %>%
    ggplot(aes(x = SSgroup.virus.avg, y = lcpm)) +
    facet_grid(. ~ tmpID, scales = "free", space = "free") +
    ylab("Log2(Counts per million)") +
    xlab("Shedding Group") +
    scale_fill_grey(start = 0.35, end = 1) + 
    geom_point(position = position_dodge(width=0.75), aes(group = group), show.legend = FALSE) +
    geom_boxplot(alpha = 0.5) +
    geom_label_repel(aes(label = bird, group = group, fill = NULL), position = position_dodge(width=0.75), show.legend=FALSE) +
    theme_classic() +
    labs(title= paste0(target, " - ", annot.target[1,2])) +
    theme(legend.title = element_blank())
    
  print(plot)
}

for(target in sort(unique(lcpm.DE$identifier))) {
  aprioriPlotting(target)
}
```


### Annotations for differentially expressed genes
ns denotes non-significant genes for each comparison and numerical values are the log(Fold change) difference
```{r message=FALSE, warning=FALSE}
tmp1a <- tmp1 %>% select(ensembl_gene_id, adj.P.Val, gene_biotype, logFC, hgnc_symbol, description) %>% mutate(comp = "LvM")
tmp2a <- tmp2 %>% select(ensembl_gene_id, adj.P.Val, gene_biotype, logFC, hgnc_symbol, description) %>% mutate(comp = "MvH")
tmp3a <- tmp3 %>% select(ensembl_gene_id, adj.P.Val, gene_biotype, logFC, hgnc_symbol, description) %>% mutate(comp = "LvH")

bind_rows(tmp1a, tmp2a, tmp3a) %>% 
  select(-description) %>% 
  filter(ensembl_gene_id %in% dt.tib$gene) %>% 
  mutate(logFoldCount = ifelse(adj.P.Val < 0.10, round(logFC, 2), "ns")) %>% 
  select(-adj.P.Val, -logFC, -gene_biotype) %>% 
  pivot_wider(names_from = comp, values_from = logFoldCount) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

#### Gene functions
```{r message=FALSE, warning=FALSE}
bind_rows(tmp1a, tmp2a, tmp3a) %>% 
  select(-comp, -adj.P.Val, -logFC) %>% 
  filter(ensembl_gene_id %in% dt.tib$gene) %>% 
  distinct(.) %>%
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```



#### Pathway analysis
```{r message=FALSE, warning=FALSE}
de.entrez <- annot %>% filter(ensembl_gene_id %in% dt.tib$gene) %>% na.omit(entrezgene_id)
entrez.universe <- annot %>% na.omit(entrezgene_id)

K <- kegga(de.entrez$entrezgene_id, universe = entrez.universe$entrezgene_id, species.KEGG = "apla") %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column("PathwayID") %>% 
  filter(P.DE < 0.05) %>% 
  arrange(P.DE)

keggList <- getGeneKEGGLinks(species.KEGG = "apla") %>% 
  as_tibble() %>% 
  rename("entrezgene_id" = GeneID)

keggGenes = data.frame()
for (path in nrow(K)){
    procPath <- filter(keggList, entrezgene_id %in% de.entrez$entrezgene_id)
    df <- data.frame(procPath)
    keggGenes <- rbind(keggGenes, df)
}

inner_join(keggGenes, K) %>%
  as_tibble() %>% 
  mutate(entrezgene_id = as.numeric(entrezgene_id)) %>% 
  left_join(annot) %>% 
  select(Pathway, PathwayID, N, DE, P.DE, ensembl_gene_id, hgnc_symbol, entrezgene_id) %>% 
  kable() %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

